---
title: "Daily Performance Stats"
subtitle: "Quick demo of the daily performance statistics function"
author: Mike Aguilar | https://www.linkedin.com/in/mike-aguilar-econ/ 
format: html
editor: visual
toc: true
toc-depth: 5
toc-location: left
embed-resources: true
execute: 
  warning: false
  echo: true
---

# Housekeeping

```{r}

#| echo: false
#| warning: false

cat("\014")  # clear console
rm(list=ls())  # Clear the workspace
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
suppressPackageStartupMessages(library(here))
source(here("Supporting", "?"))

```

# Data

Task: set seed 124 for simulations

```{r}
 set.seed?
```

Task: Create a dataframe called price with Date, Asset, and Benchmark columns. Dates should be daily starting from 2008-01-01 for 5yrs of daily trading activity. Asset and Benchmark should be cumulative products of random normal returns.

```{r}
 n <- 250*?
 Price <- data.frame(
   Date      = seq.Date(from = as.Date("2008-01-01"), by = "day", length.out = n),
   Asset     = ?(1 + rnorm(n, 0.0005, 0.02)),
   Benchmark = ?(1 + rnorm(n, 0.0003, 0.015))
 )
 
```

# Peformance Stats

Task: Set meta data

```{r}
 Meta <- list(asset_name = "MyPortfolio", benchmark_name = "Bencmark")
```

Task: Call DailyPerformanceStats function with Price, Meta, rf_annual = 0, and output_dir set to "../Output/DailyPerformanceStat"

```{r}
 results <- DailyPerformanceStats(?, Meta, rf_annual = 0, output_dir ="./Output/DailyPerformanceStat")
```

Task: Display sample_stats_tbl

```{r}
?$sample_stats_tbl
```

Q: Interpret the results.

A:




Task: Display the risk_return_tbl

```{r}
results$?
```

Q: Interpret the results.

A:




Task: Display the drawdowns_tbl

```{r}
results$?
```

Q: Interpret the results. Opening the MyPortfolio_underwater.png might be helpful.

A:




Task: Display all of the plots

```{r}
?

```
