---
title: "Clustering Momentum"
subtitle: "Using GMM soft clustering to build a momentum signal"
author: Mike Aguilar | https://www.linkedin.com/in/mike-aguilar-econ/ 
format: html
editor: visual
toc: true
toc-depth: 5
toc-location: left
embed-resources: true
execute: 
  warning: false
  echo: true
---

# Background

A slight variation of the classic 12-2 momentum signal has the following characteristics

-   Long assets with high 12mth returns and low 2mth returns
-   Short stocks with low 12mth returns and high 2mth returns
-   Neutral all other stocks

Our empirical goal is to cluster stocks by the 12mth and 2mth returns so that we can construct the 12-2 momentum signal.

In this exercise we will use soft clustering via Gaussian Mixture Models (GMM).

# Housekeeping

```{r}
#| echo: false
cat("\014")  # clear console
rm(list=ls())  # Clear the workspace
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
suppressPackageStartupMessages(library(here))
source(here("Supporting", "PackageLoads.R"))
```

# Data

## Load simple returns

Task: Load SP500-Monthly Simple Returns

```{r}
Returns <- ?(here("Data","SP500_Mthly_SimpleReturns.csv"))
```

## Load monthly momentum

```{r}
Momentum <- ?(here("Data","SP500_Mthly_Momentum.csv"))
```

## Merge

Task: Merge the returns and momentum, retaining only those dates with complete information

```{r}
merged_df <- Returns %>%
  left_join(Momentum, ? = c("Date", "ticker"))
```

## Set Test and Train Periods

Task: Create a dataframe for the Train period which spans the beginning of the dataset to Aug2025. Call this dataframe Test. Create a dataframe for the Train period which is Sep2025 period. Call this dataframe Train.

```{r}
test_date <- as.Date("2025-09-30") 
last_train_date <- as.Date(
  format(test_date, "%Y-%m-01")  ) - 1

Train <- merged_df %>% filter(Date < test_date)
Train.Returns  <- Train %>% ?(!momentum)
Train.Momentum <- Train %>% ?(!OneMthSimpleRet)

Test <- merged_df %>% ?(Date == test_date)
Test.Returns  <- Test %>% ?(!momentum)
Test.Momentum <- Test %>% ?(!OneMthSimpleRet)

```

## Isolate Index and Constituents

Task: Isolate the returns for the index and the group of constituents into separate dataframes.

```{r}
Train.Constituents <- Train %>%
  filter(!ticker == "SP500")
Train.Returns.Index <- Train.Returns %>%
  filter(ticker == ?)
Train.Returns.Constituents <- Train.Returns %>%
  filter(?) == "SP500")

Test.Constituents <- Test %>%
  filter(> == "SP500")
Test.Returns.Index <- Test.Returns %>%
  filter(? == "SP500")
Test.Returns.Constituents <- Test.Returns %>%
  filter(? == "SP500")

```

# Signal (Static)

## Explore number of clusters

Task: use the m/m simple returns to compute 12mth and 2mth returns.

```{r}
Train.Returns.Constituents <- Train.Returns.Constituents %>%
  arrange(ticker, Date) %>%
  group_by(ticker) %>%
  mutate(
    Ret12m =
      (1 + lag(OneMthSimpleRet,  0)) *
      (1 + lag(OneMthSimpleRet,  ?)) *
      (1 + lag(OneMthSimpleRet,  2)) *
      (1 + lag(OneMthSimpleRet,  3)) *
      (1 + lag(OneMthSimpleRet,  4)) *
      (1 + lag(OneMthSimpleRet,  5)) *
      (1 + lag(OneMthSimpleRet,  6)) *
      (1 + lag(OneMthSimpleRet,  7)) *
      (1 + lag(OneMthSimpleRet,  8)) *
      (1 + lag(OneMthSimpleRet,  9)) *
      (1 + lag(OneMthSimpleRet, 10)) *
      (1 + lag(OneMthSimpleRet, 11)) *
      (1 + lag(OneMthSimpleRet, 12)) ?,
    Ret2m =
      (1 + lag(OneMthSimpleRet,  0)) *
      ?,
    
  ) %>%
  ungroup()
```

Task: Create a temporary df that contains the returns of the constituents for the most recent training period.

```{r}
df <- Train.Returns.Constituents %>%
  filter(Date == ?(?, na.rm = TRUE)) %>%
  na.omit()
```

Task: select only the 12mth and 2mth returns and stdz them

```{r}
X <- as.matrix(df %>% select(Ret12m, Ret2m))
X_scaled <- ?(X)  
```

Task: Loop over a grid of K=1:10 potential clusters, use MClust on each iteration of the loop, and compute the loglikelihood, AIC, and BIC for each model

```{r}
n <- nrow(X)
K_grid <- 1:10
gmm_ic_tbl <- lapply(K_grid, function(k) {

  m <- Mclust(data = X_scaled, G = k)
  ll <- as.numeric(logLik(m))
  p  <- attr(logLik(m), "df")

  tibble(
    k = k,
    logLik = ll,
    params = p,
    AIC = -2*ll + 2*p,
    BIC = -2*ll + log(n)*p,
  )
}) %>% bind_rows()

gmm_ic_tbl
```

Q: What is the implication of the table?

A: 

## Cluster

Task: Form 3 clusters using GMM via Mclust

```{r}
cluster_gmm3 <- ?(data = ?, G = ?)
```

Task: Add cluster_id to your df via the cluster_gmm3 classification. Also construct a score as the (1+Ret12m)/(1+Ret2m)-1

```{r}
df <- df %>%
  mutate(
    cluster_id = cluster_gmm3$?,
    score = ?
  )
```

Task: Compute cluster-level average score

```{r}
cluster_stats <- df %>%
  group_by(cluster_id) %>%
  summarise(
    mu_12m = ?(Ret12m, na.rm = TRUE),
    mu_2m  = ?(Ret2m,  na.rm = TRUE),
    mean_score = mean(score,  na.rm = TRUE),
    .groups = "drop"
  )
```

Task: identify Long / Short clusters

```{r}
long_cluster<- cluster_stats %>% slice_?(mean_score, n = 1) %>% pull(?)
long_cluster
short_cluster<- cluster_stats %>% slice_?(mean_score, n = 1) %>% pull(?)
short_cluster
```

Task: Add long/short labels to each asset

```{r}
df <- df %>%
  mutate(cluster_label = case_when(
    cluster_id == ?  ~ "Long",
    cluster_id == ? ~ "Short",
    TRUE                     ~ "Neutral"
  ))
df
```

Task: add cluster labels to add each cluster

```{r}
centroids_labeled <- cluster_stats %>%
  mutate(
    cluster_id = as.integer(cluster_id),
    cluster_label = case_when(
      cluster_id == as.integer(long_cluster)  ~ "Long",
      cluster_id == as.integer(short_cluster) ~ "Short",
      TRUE                                    ~ "Neutral"
    )
  )
```

Task: Plot the clusters

```{r}
ggplot(df %>% mutate(cluster_id = factor(?)),
       aes(x = Ret2m, y = Ret12m, color = cluster_label)) +
  geom_point(alpha = 0.75, size = 2) +
  geom_point(
    data = centroids_labeled,
    aes(x = mu_2m, y = mu_12m),
    inherit.aes = FALSE,
    size = 4
  ) +
  geom_text(
    data = centroids_labeled,
    aes(x = mu_2m, y = mu_12m, label = cluster_label),   # <-- add label
    inherit.aes = FALSE,
    vjust = -1,
    show.legend = FALSE
  ) + 
  labs(
    color = "GMM cluster",
    x = "2-month return",
    y = "12-month trailing return",
    title = "GMM (k=3) clusters labeled by momentum score"
  ) +
  theme_minimal()
```

## Construct the Signal

Task: Assign cluster probabilities to conform to the 12-2 momentum signal. Store these probabilities within a df called DegreeBelonging with rows = assets and columns = clusters, with each entry being the degree of belonging.

Task: Gather the posterior probabilities

```{r}
Z <- cluster_gmm3$?
stopifnot(nrow(Z) == nrow(df))
head(Z)
```

Q: Interpret the first row.

A: 

Task: Rank clusters by mean momentum score

```{r}
cluster_rank <- df %>%
  ?(cluster_id) %>%
  summarise(mean_score = ?(score, na.rm = TRUE), .groups = "drop") %>%
  arrange(?)
```

Task: Cluster ids ordered from low-\>high score

```{r}
ord_ids <- cluster_rank$cluster_id
short_id <- ord_ids[?]
long_id  <- ord_ids[?]
neutral_id <- ord_ids[?]
```

Task: Build DegreeBelonging: reorder probability columns to (Short, Neutral, Long) Note: columns of Z are labeled 1:K (component numbers)

```{r}
DegreeBelonging <- as.data.frame(Z[, c(short_id, neutral_id, long_id), drop = FALSE])
colnames(DegreeBelonging) <- c("Short", "Neutral", "Long")
rownames(DegreeBelonging) <- df$ticker

DegreeBelonging
```

Task: Create a scatter plot illustrating the potentially overlapping clusters.

```{r}
df_overlap <- df %>%
  mutate(
    cluster_id = as.integer(cluster_gmm3$classification),
    max_prob   = apply(Z, 1, max),
   cluster_label = df$cluster_label
  )

ggplot(?, aes(x = Ret2m, y = Ret12m)) +
  geom_point(aes(color = cluster_label, alpha = 1 - max_prob), size = 2) +
  scale_alpha_continuous(range = c(0.1, 1), guide = "none") +
  labs(
    x = "2-month return",
    y = "12-month return",
    title = "Overlapping clusters: higher transparency = higher uncertainty (lower max posterior)"
  ) +
  theme_minimal()
```

Task: Cull ambiguous assets. Drop any assets whose max cluster probability less than "confidence threshold", which you can set to 60%.

```{r}
conf_thresh <- ? 
keep_idx <- apply(Z, 1, max) > conf_thresh


df_culled <- df [keep_idx, ]
df_culled
```

Task: Assign clusters. For each of the remaining assets assign the cluster with the highest probability.

```{r}
cluster_map <- max.col(Z, ties.method = "first")   # returns 1..K
cluster_assigned <- ?[keep_idx]

df_assigned <- df[keep_idx, ] %>%
  mutate(cluster_id = as.integer(cluster_assigned))

df_assigned
```

Task: Create a scatter plot illustrating to which cluster each asset is defined.

```{r}
ggplot(df_assigned, aes(x = Ret2m, y = Ret12m, color = cluster_label)) +
  geom_point(alpha = 0.7) +
  geom_point(
    data = centroids_labeled ,
    aes(x = mu_2m, y = mu_12m),
    inherit.aes = FALSE, size = 4, stroke = 1.5
  ) +
  geom_text(
    data = centroids_labeled ,
    aes(x = mu_2m, y = mu_12m, label = cluster_label),
    inherit.aes = FALSE, vjust = -1, show.legend = FALSE
  ) +
  labs(
    color = "GMM cluster",
    x = "2-month return",
    y = "12-month trailing return",
    title = "GMM (k=3) clusters labeled by momentum score"
  )
```

Task: Assign signal value to each asset. Your output should be a df where each row is an asset, the column is the position, and each entry is either +1, 0, or -1.

```{r}
SignalStatic <- df %>%
  transmute(
    ticker,
    signal_position = case_when(
      cluster_label == "Long"  ~  1L,
      cluster_label == "Short" ~ -1L,
      TRUE             ~  0L
    )
  )

SignalStatic
```

## Test the Signal

Task: Call out to signal static to evaluate the efficacy of this signal.

```{r}
Meta <- list(assetname = "asset", benchmarkname = "market", signalname = "GMM3Momentum")
  
signal_position <- data.frame(ticker = SignalStatic$ticker, 
                              signal_position = SignalStatic$signal_position)

test_returns <- data.frame(ticker = Test.Returns.Constituents$ticker, 
                           test_returns = Test.Returns.Constituents$OneMthSimpleRet)

return_threshold <- 0.05
SignalEval <- ?(
    test_returns = test_returns,
    Meta = Meta,
    return_threshold = return_threshold,
    signal_position = signal_position
  )
SignalEval
```

# Signal (Dynamic)

## Restructure Test/Train

Task: Roll your test/train samples from 2017-01-31 through Nov2025

```{r}
# split data
test_begin_date <- as.Date("2017-01-31") 
last_train_date <- as.Date(
  format(test_date, "%Y-%m-01")  ) - 1

Train <- merged_df %>% filter(Date < test_date)
Train.Returns  <- Train %>% select(!momentum)
Train.Momentum <- Train %>% select(!OneMthSimpleRet)

Test <- merged_df %>% filter(Date >= test_begin_date) 
Test.Returns  <- Test %>% select(!momentum)
Test.Momentum <- Test %>% select(!OneMthSimpleRet)
```

Task: Isolate the returns for the index and the group of constituents into separate dataframes.
```{r}
Train.Constituents <- Train %>%
  filter(!ticker == "SP500")
Train.Returns.Index <- Train.Returns %>%
  filter(ticker == "SP500")
Train.Returns.Constituents <- Train.Returns %>%
  filter(!ticker == "SP500")
Train.Momentum.Index <- Train.Momentum %>%
  filter(ticker == "SP500")
Train.Momentum.Constituents <- Train.Momentum %>%
  filter(!ticker == "SP500")

Test.Constituents <- Test %>%
  filter(!ticker == "SP500")
Test.Returns.Index <- Test.Returns %>%
  filter(ticker == "SP500")
Test.Returns.Constituents <- Test.Returns %>%
  filter(!ticker == "SP500")
Test.Momentum.Index <- Test.Momentum %>%
  filter(ticker == "SP500")
Test.Momentum.Constituents <- Test.Momentum %>%
  filter(!ticker == "SP500")
```


## Cluster

Task: Let's use K=3 clusters to align with signal construction. Repeat the GMM clustering for each test period. Output should be a long df with cols = (time, assets,cluster 1 prob, cluster 2 prob, cluster 3 prob)

```{r}
# 12 month and 2 month return
Train.Returns.Constituents <- Train.Returns.Constituents %>%
  arrange(ticker, Date) %>%
  group_by(ticker) %>%
  mutate(
    Ret12m =
      (1 + lag(OneMthSimpleRet,  0)) *
      (1 + lag(OneMthSimpleRet,  1)) *
      (1 + lag(OneMthSimpleRet,  2)) *
      (1 + lag(OneMthSimpleRet,  3)) *
      (1 + lag(OneMthSimpleRet,  4)) *
      (1 + lag(OneMthSimpleRet,  5)) *
      (1 + lag(OneMthSimpleRet,  6)) *
      (1 + lag(OneMthSimpleRet,  7)) *
      (1 + lag(OneMthSimpleRet,  8)) *
      (1 + lag(OneMthSimpleRet,  9)) *
      (1 + lag(OneMthSimpleRet, 10)) *
      (1 + lag(OneMthSimpleRet, 11)) *
      (1 + lag(OneMthSimpleRet, ?)) - ?,
    Ret2m =
?, 
    
  ) %>%
  ungroup() %>%
  na.omit()
```

Task: Main loop over each time period in the training set to fit GMM and extract posterior probabilities
```{r}
set.seed(1)
next_month <- function(d) ceiling_date(as.Date(d), "month") %m+% months(1)-days(1)


K <- 3L
relabel_probs_by_score <- function(Z, score_vec) {
  mean_score <- sapply(1:ncol(Z), function(g) sum(Z[, g] * score_vec) / sum(Z[, g]))
  ord <- order(mean_score)  # low -> high
  Z2 <- Z[, ord, drop = FALSE]
  colnames(Z2) <- c("p_Short", "p_Neutral", "p_Long")
  Z2
}


# Training period
dates_vec <- sort(unique(as.Date(Train.Returns.Constituents$Date)))

# main loop
gmm_probs_labeled_long <- map_dfr(dates_vec, function(date_i) {

  df_t <- Train.Returns.Constituents %>%   
    filter(as.Date(Date) == as.Date(date_i)) %>%
    transmute(
      time  = as.Date(date_i),
      asset = ticker,
      Ret12m = as.numeric(Ret12m),
      Ret2m  = as.numeric(Ret2m)
    ) %>%
    filter(is.finite(Ret12m), is.finite(Ret2m)) %>%
    distinct(time, asset, .keep_all = TRUE) %>%
    mutate(score = (1 + Ret12m) / (1 + Ret2m) - 1) %>%
    filter(is.finite(score))

  if (nrow(df_t) < K) {
    return(tibble(
      time = df_t$time,
      asset = df_t$asset,
      p_Short = NA_real_,
      p_Neutral = NA_real_,
      p_Long = NA_real_
    ))
  }


  # fit GMM
  X <- as.matrix(df_t %>% select(Ret12m, Ret2m))
  X_scaled <- scale(X)
  
  fit <- tryCatch(
    suppressMessages(suppressWarnings(
      Mclust(X_scaled, G = K, verbose = FALSE)
    )),
    error = function(e) NULL
  )

  # If fit failed, return NA probs
  if (is.null(fit) || is.null(fit$z) || ncol(fit$z) != K) {
    return(tibble(
      time = df_t$time,
      asset = df_t$asset,
      p_Short = NA_real_,
      p_Neutral = NA_real_,
      p_Long = NA_real_
    ))
  }

  # Relabel posterior probs to Short/Neutral/Long each period
  Z <- fit$z
  Z_labeled <- relabel_probs_by_score(Z, df_t$score)

  # Output long df for this period
  tibble(
    time = as.Date(next_month(date_i)),
    asset = df_t$asset,
    p_Short   = Z_labeled[, "p_Short"],
    p_Neutral = Z_labeled[, "p_Neutral"],
    p_Long    = Z_labeled[, "p_Long"]
  )
})

# Result
gmm_probs_labeled_long
```

Task: Cull ambiguous assets. Drop any assets whose max cluster probability less than "confidence threshold", which you can set to 60%.

```{r}
conf_thresh <- 0.60
Z <- as.matrix(gmm_probs_labeled_long[, c("p_Short", "p_Neutral", "p_Long")])

# keep/drop index
max_prob <- apply(Z, 1, max)
keep_idx <- max_prob >= conf_thresh

# culled dataset (keep only confident asset×time rows)
gmm_probs_culled <- gmm_probs_labeled_long[keep_idx, ] %>%
  mutate(max_prob = max_prob[keep_idx])
gmm_probs_culled
```

Task: Assign clusters. For each of the remaining assets assign the cluster with the highest probability.

```{r}
Zc_culled <- as.matrix(gmm_probs_culled[, c("p_Short", "p_Neutral", "p_Long")])

gmm_assigned_culled <- gmm_probs_culled %>%
  mutate(
    cluster_label = c("Short", "Neutral", "Long")[max.col(Zc_culled, ties.method = "first")]
  ) %>%
  select(-p_Short, -p_Neutral, -p_Long)

gmm_assigned_culled
```

## Construct the Signal

Task: Assign signal value to each asset. Your output should be a df where cols = (time, asset, position), where signal position {+1, 0, or -1}.

```{r}
Zc <- as.matrix(gmm_probs_labeled_long[, c("p_Short", "p_Neutral", "p_Long")])

gmm_assigned <- gmm_probs_labeled_long %>%
  mutate(
    cluster_label = c("Short", "Neutral", "Long")[max.col(Zc, ties.method = "first")]
  ) %>%
  select(-p_Short, -p_Neutral, -p_Long)

gmm_assigned
```

Task: Form the signal matrix

```{r}
SignalDynamic <- gmm_assigned %>%
  transmute(
    time,
    asset,
    position = case_when(
      cluster_label == "Long"  ~  1L,
      cluster_label == "Short" ~ -1L,
      TRUE             ~  0L
    )
  )

SignalDynamic
```

## Test the Signal

Task: Call out to signal evaluation dynamic to evaluate the efficacy of this signal.

```{r}
returns_panel <- Test.Returns.Constituents %>% 
  filter(Date %in% unique(SignalDynamic$time)) %>%
  transmute(ticker, date = Date, test_returns = OneMthSimpleRet)

signal_position_panel <- SignalDynamic %>% 
  transmute(ticker = asset, date = time, signal_position = position)

Meta <- list(assetname = "Universe", benchmarkname = "SP500", signalname = "GMMMomentum")

res_dyn <- ?(
  signal_position_ts = signal_position_panel,
  returns_ts = returns_panel,
  Meta = Meta,
  return_threshold = 0.05,    # direction mode effectively in static
  keep_static_results = FALSE,
  verbose = TRUE
)
```
