---
title: "Clustering Momentum"
subtitle: "Using Hierarchical clustering to build a momentum signal"
author: Mike Aguilar | https://www.linkedin.com/in/mike-aguilar-econ/ 
format: html
editor: visual
toc: true
toc-depth: 5
toc-location: left
embed-resources: true
execute: 
  warning: false
  echo: true
---

# Background

Let's explore the classic 12-2 momentum signal with the following characteristics

-   Long assets with high momentum
-   Short stocks with low momentum
-   Neutral all other stocks

Our empirical goal is cluster stocks by the momentum so that we can construct the 12-2 momentum signal.

# Housekeeping

```{r}
#| echo: false
cat("\014")  # clear console
rm(list=ls())  # Clear the workspace
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
suppressPackageStartupMessages(library(here))
source(here("Supporting", "PackageLoads.R"))
```

# Data

## Load simple returns

Task: Load SP500-Monthly Simple Returns

```{r}
Returns <- read.csv(here("?","SP500_Mthly_SimpleReturns.csv"))
```

## Load monthly momentum

```{r}
Momentum <- read.csv(here("?","SP500_Mthly_Momentum.csv"))
```

## Merge

Task: Merge the returns and momentum, retaining only those dates with complete information

```{r}
merged_df <- Returns %>%
  left_join(?, by = c("Date", "ticker"))

```

## Set Test and Train Periods

Task: Create a dataframe for the Train period which spans the beginning of the dataset to Aug2025. Call this dataframe Test. Create a dataframe for the Train period which is Sep2025 period. Call this dataframe Train.

```{r}
test_date <- as.Date("2025-09-30") 
last_train_date <- as.Date(
  format(test_date, "%Y-%m-01")  ) - 1

Train <- merged_df %>% filter(Date < test_date)
Train.Returns  <- Train %>% select(!momentum)
Train.Momentum <- Train %>% select(!OneMthSimpleRet)

Test <- merged_df %>% filter(Date == test_date)
Test.Returns  <- Test %>% select(!momentum)
Test.Momentum <- Test %>% select(!OneMthSimpleRet)

```

## Isolate Index and Constituents

Task: Isolate the returns for the index and the group of constituents into separate dataframes. Repeat for momentum.

```{r}
Train.Constituents <- Train %>%
  filter(!ticker == "SP500")
Train.Returns.Index <- Train.Returns %>%
  filter(ticker == "SP500")
Train.Returns.Constituents <- Train.Returns %>%
  filter(!ticker == "SP500")
Train.Momentum.Index <- Train.Momentum %>%
  filter(ticker == "SP500")
Train.Momentum.Constituents <- Train.Momentum %>%
  filter(!ticker == "SP500")

Test.Constituents <- Test %>%
  filter(!ticker == "SP500")
Test.Returns.Index <- Test.Returns %>%
  filter(ticker == "SP500")
Test.Returns.Constituents <- Test.Returns %>%
  filter(!ticker == "SP500")
Test.Momentum.Index <- Test.Momentum %>%
  filter(ticker == "SP500")
Test.Momentum.Constituents <- Test.Momentum %>%
  filter(!ticker == "SP500")

```

# Hierarchical Clustering

```{r}
mostrecenttrain<-Train.Momentum.Constituents %>%
    filter(Date == ?(Date, na.rm = TRUE)) %>%
   na.omit()
```

```{r}
X <- matrix(mostrecenttrain$momentum, ncol = 1)
rownames(X) <-mostrecenttrain$ticker
head(X)
```

Task: Compute the euclidean distance between each asset's momentum

```{r}
D  <- dist(X, method = "?")
head(D)
```

Q: We could have also chosen other distance methods, such as Manhattan. Which might be more sensitive to market crashes, Euclidean or Manhattan?

A: 

Task: Form an agglomerative tree using Ward.D2 method based on the hclust function

```{r}
hc <- ?(D, method = "ward.D2")
```


Task: Plot the associated dendrogram

```{r}
plot(?, labels = FALSE, hang = -1,
     main = "Dendrogram Tree (Ward.D2, Euclidean) — Momentum",
     xlab = "", sub = "")
```

Q: Interpret the dendrogram

A:



Task: Impose 3 clusters by "cutting" the tree using cutree. We choose 3 because we'll soon need Long, Neutral, Short cluster.

```{r}
clusters <- ?(hc, k = ?)
head(clusters)
```

Task: Plot the dendogram and overlay 3 clusters using the rect.hclust command

```{r}
plot(hc, labels = FALSE, hang = -1,
     main = paste("Dendrogram Tree (Ward.D2, Euclidean)", "cut at K =", 3),
     xlab = "", sub = "")

?(?, k = ?, border = 2:6)
```

Task: Mutate mostrecenttrain to include a column with cluster ID

```{r}
mostrecenttrain$clusterID <- clusters[ mostrecenttrain$?]

```

# Signal Construction

Task: Find the average momentum for each cluster

```{r}
ClusterStrength<-mostrecenttrain %>%
  group_by(clusterID) %>%
  summarize(
    AvgMomentum = ?(momentum)
  )
ClusterStrength
```

Task: identify Long / Short clusters

```{r}
long_cluster  <- ClusterStrength %>%
  slice_?(AvgMomentum, n = 1, with_ties = TRUE) %>%
  pull(?) %>%
  unique()

short_cluster <- ClusterStrength %>%
  slice_?(AvgMomentum, n = 1, with_ties = TRUE) %>%
  pull(?) %>%
  unique()
```

Task: Add cluster labels to the mostrecenttrain df to form the signal_position column

```{r}
 mostrecenttrain <- mostrecenttrain %>%
  mutate(clusterID = as.integer(clusterID)) %>%
  mutate(
    signal_position = case_when(
      clusterID %in% as.integer(?)  ~  1L,
      clusterID %in% as.integer(?) ~ -1L,
      TRUE                                     ~  0L   # neutral / no position
    )
  )
```

Task: prep the df for evaluation

```{r}
Meta <- list(assetname = "asset", benchmarkname = "market", signalname = "K4HierMomentum")
  
signal_position <- data.frame(ticker = mostrecenttrain$ticker, 
                              signal_position = mostrecenttrain$signal_position)

test_returns <- data.frame(ticker = Test.Returns.Constituents$ticker, 
                           test_returns = ?)

```

Task: evaluate the signal with a return threshold of 0.05

```{r}
return_threshold <- ?
SignalEval <- ?(
    test_returns = ?,
    Meta = Meta,
    return_threshold = return_threshold,
    signal_position = signal_position
  )
SignalEval
```
