---
title: "Momentum Portfolio"
subtitle: "Combining two strategies to form a portfolio"
author: Mike Aguilar | https://www.linkedin.com/in/mike-aguilar-econ/ 
format: html
editor: visual
toc: true
toc-depth: 5
toc-location: left
embed-resources: true
execute: 
  warning: false
  echo: true
---

# Housekeeping

```{r}
#| echo: false
cat("\014")  # clear console
rm(list=ls())  # Clear the workspace
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
suppressPackageStartupMessages(library(here))
source(here("Supporting", "PackageLoads.R"))

```

# Data

## Load monthly momentum

```{r}
Momentum <- read.csv(here("Data","SP500_Mthly_Momentum.csv"))
```

## Load monthly returns

```{r}
SP500_Mthly_Returns <- read.csv(here("Data","SP500_Mthly_SimpleReturns.csv"))
```

## Daily Data for Evaluation

Task: Load SP500 Daily Simple Returns

```{r}
SP500_Daily_SimpleReturns <- read.csv(here("Data","SP500_Daily_SimpleReturns.csv"))
```

Task: Ensure Date is Date class

```{r}
SP500_Daily_SimpleReturns <- SP500_Daily_SimpleReturns %>%
  mutate(Date = as.Date(Date))
```

Task: Grab the names of the index constituents

```{r}
constituent_names <- unique(SP500_Daily_SimpleReturns$ticker)
constituent_names <- constituent_names[constituent_names != "SP500"]
```

## Set Train, Evaluation, Test Periods

Task: Set the train period to be all dates prior to 09-30-2025. Set the evaluation period the month of September 2025. Set the test period to be October 2025.

```{r}
test_end_date <- as.Date("2025-10-31")
evaluation_end_date <- as.Date("2025-09-30")
train_end_date <- as.Date(
  format(evaluation_end_date, "%Y-%m-01")  ) - 1
```

Task: Create a dataframe called Train.Momentum that captures momentum for all the constituents on all dates prior to the test date of 2025-09-30. Also create an analogous df for returns.

```{r}

Train.Momentum.Constituents <- Momentum %>% 
  filter(Date<=train_end_date) %>%
  filter(!(ticker == "SP500"))

Train.Returns.Constituents <- SP500_Mthly_Returns %>%
  filter(Date<=train_end_date) %>%
  filter(!(ticker == "SP500"))

```

Task: Create a dataframe for the evaluation period returns

```{r}

Evaluation.Returns.Constituents <- SP500_Daily_SimpleReturns %>%
  filter(
    Date > train_end_date &
    Date <= evaluation_end_date &
    ticker != "SP500"
  )


```

Task: Create a dataframe for the test period returns

```{r}
Test.Returns.Constituents <- SP500_Daily_SimpleReturns %>%
  filter(
    Date > evaluation_end_date &
    Date <= test_end_date & 
    ticker != "SP500"
  ) 
```

# Momentum Signal

Task: Construct a simple momentum strategy

-   Define LongCondition = 1 if $mom^{Train}_{i} >0$ and 0 otherwise

-   Define ShortCondition = 1 if $mom^{Train}_{i} < 0$ and 0 otherwise

-   $s_{i,t}$ = LongCondition minus ShortCondition

Task: Create a dataframe that contains the momentum on the last observation of the training period

```{r}
Signal_Matrix <- Train.Momentum.Constituents %>% 
  filter(Date == train_end_date) %>%
  select(Date, ticker, momentum) %>% 
  rename("signal_matrix" = "momentum")

Signal_Matrix
```

## Strategy Weights

Task: Create a weights dataframe based upon the signal. Set the weight equal to the signal value

```{r}
Weights.Momentum.Raw <- Signal_Matrix %>% 
  transmute(ticker, Date, weight = signal_matrix)
```

Task: Scale weights so that fully invested (sum of abs(weights) = 1)

```{r}
gross_scale <- sum(abs(Signal_Matrix$signal_matrix), na.rm = TRUE)
gross_scale

Weights.Momentum <- Weights.Momentum.Raw %>%
  mutate(weight = Signal_Matrix$signal_matrix / gross_scale)

sum(abs(Weights.Momentum$weight))

```

Task: Construct a strategy for the Evaluation period by calling out to construct_portfolio

```{r}
Strategy.Momentum <- construct_portfolio(
    Evaluation.Returns.Constituents,
    Weights.Momentum,
    # column names in inputs
    date_col   = "Date",
    ticker_col = "ticker",
    return_col = "OneDaySimpleRet",
    weight_col = "weight",
    return_type = "simple", #c("simple","log"),
    rebalance_freq = "monthly", #c("monthly","quarterly","yearly"),
    shift_n = 1 # if shift_n=1 and rebalance_freq=monthly, then w_Jan * ret_Feb
)
```

# High SR Signal

Task: Compute the Sharpe Ratio for each stock in the training period.

```{r}
SR_Matrix <- Train.Returns.Constituents %>%
  filter(!(ticker == "SP500")) %>%
  group_by(ticker) %>%
  summarize(
    mean_return = mean(OneMthSimpleRet, na.rm = TRUE),
    sd_return = sd(OneMthSimpleRet, na.rm = TRUE)
  ) %>%
  mutate(
    sharpe_ratio = mean_return / sd_return
  ) %>%
  select(ticker, sharpe_ratio)
```

Task: Create a signal matrix based upon Sharpe Ratio

```{r}
SR_Signal_Matrix <- SR_Matrix %>%
  mutate(
    signal_matrix = ifelse(sharpe_ratio > 0, sharpe_ratio, 0),
    Date = train_end_date
  ) %>%
  select(ticker, signal_matrix, Date)
```

Let's use these Sharpe Ratio values as the basis for weights for the second strategy. Let's also scale them so that fully invested.

## Strategy Weight

Task: Scale weights so that fully invested

```{r}
gross_scale <- sum(SR_Signal_Matrix$signal_matrix)
Weights.SR <- SR_Signal_Matrix %>%
  mutate(weight = SR_Signal_Matrix$signal_matrix / gross_scale) %>%
  rename("ticker" = "ticker","weight" = "weight")
sum(Weights.SR$weight)
```

Task: Construct the strategy for the evaluation period by calling out to construct_portfolio

```{r}
Strategy.SR <- construct_portfolio(
    Evaluation.Returns.Constituents,
    Weights.SR,
    # column names in inputs
    date_col   = "Date",
    ticker_col = "ticker",
    return_col = "OneDaySimpleRet",
    weight_col = "weight",
    return_type = "simple", #c("simple","log"),
    rebalance_freq = "monthly", #c("monthly","quarterly","yearly"),
    shift_n = 1 # if shift_n=1 and rebalance_freq=monthly, then w_Jan * ret_Feb
)

```

# Portfolio

Let's combine the two strategies to form a portfolio

## User Inputs

### Portfolio Returns and Cov

Task: create a returns_matrix with momentum returns and SR returns

```{r}
temp1 <- Strategy.Momentum$portfolio %>% select(date, pnl)
temp2 <- Strategy.SR$portfolio       %>% select(date, pnl)

# join by date to ensure alignment
returns_df <- temp1 %>%
  left_join(temp2, by = "date", suffix = c("_Momentum", "_SR")) %>%
  rename(Date = date,
         Momentum_Return = pnl_Momentum,
         SR_Return       = pnl_SR)

# Useful for optimization
returns_matrix <- as.matrix(returns_df %>% select(Momentum_Return, SR_Return))
colnames(returns_matrix) <- c("Momentum", "SR")
m <- ncol(returns_matrix)  # should be 2

```

Task: Set expected returns and cov matrix for strategies = sample moments

```{r}
mu_s <- colMeans(returns_matrix, na.rm = TRUE)   # length m
Sigma_s <- cov(returns_matrix, use = "pairwise.complete.obs")  # m x m
```

### Portfolio Targets

Task: Set the portfolio target return to be 1% greater than the max of the two strategy mean returns

```{r}
target <- max(colMeans(returns_matrix))*1.01 
meq <- 1          # set to 1 if you want mu'w >= target (inequality); meq = 2 for exact target
```

### Asset Exposure Constraints

Task: Set the maximum asset weight to 25% and minimum to -10%

```{r}
# per-asset limits
max_asset <- .25
min_asset <- -.1     

# optional strategy box constraints 
w_min_strategy <- c(-2,-2) #rep(-1, m)  
w_max_strategy <- c(3,3) #rep(1, m)   
```

## Portfolio Optimization

Task: Conduct the Min variance portfolio optimization using quadprog, assuming fully invested along with the constraints set above.

```{r}
# objective (quadprog requires Dmat positive definite)
eps <- 1e-10
Dmat <- 2 * Sigma_s + eps * diag(m)
dvec <- rep(0, m)

# build asset exposure mapping a_j (rows = strategies, cols = assets)

w1 <- Weights.Momentum %>% select(ticker, weight) %>% rename(weight1 = weight)
w2 <- Weights.SR %>% select(ticker, weight) %>% rename(weight2 = weight)

# join into a wide matrix of asset weights by strategy
asset_wide <- full_join(w1, w2, by = "ticker") %>%
  mutate(weight1 = ifelse(is.na(weight1), 0, weight1),
         weight2 = ifelse(is.na(weight2), 0, weight2))


# asset list
assets <- asset_wide$ticker
n_assets <- length(assets)

# Build A_asset matrix of dimension (m x n_assets): each column j is a_j vector (length m)
A_asset <- rbind(asset_wide$weight1, asset_wide$weight2)  # m x n_assets
colnames(A_asset) <- assets
rownames(A_asset) <- colnames(returns_matrix)  # "Momentum","SR"


# Build Amat columns (each column is a constraint vector length m)
Amat_cols <- list()

# 1) sum(w) = 1
Amat_cols[[length(Amat_cols)+1]] <- rep(1, m)
bvec <- 1

# 2) mu_s' w = target  (equality if meq=2)
Amat_cols[[length(Amat_cols)+1]] <- mu_s
bvec <- c(bvec, target)

# 3) asset-level lower bounds: a_j^T w >= min_asset
for (j in seq_len(n_assets)) {
  Aj <- A_asset[, j]         # length m
  Amat_cols[[length(Amat_cols)+1]] <- Aj
  bvec <- c(bvec, min_asset)
}

# 4) asset-level upper bounds: -a_j^T w >= -max_asset  -> a_j^T w <= max_asset
for (j in seq_len(n_assets)) {
  Aj <- -A_asset[, j]
  Amat_cols[[length(Amat_cols)+1]] <- Aj
  bvec <- c(bvec, -max_asset)
}

# 5) strategy box constraints: w_k >= w_min_strategy[k]
for (k in seq_len(m)) {
  ek <- rep(0, m); ek[k] <- 1
  Amat_cols[[length(Amat_cols)+1]] <- ek
  bvec <- c(bvec, w_min_strategy[k])
}
# and w_k <= w_max_strategy[k]  -> (-e_k)^T w >= -w_max_strategy[k]
for (k in seq_len(m)) {
  ek <- rep(0, m); ek[k] <- -1
  Amat_cols[[length(Amat_cols)+1]] <- ek
  bvec <- c(bvec, -w_max_strategy[k])
}

# Combine to matrix: quadprog expects Amat with columns = constraint vectors
Amat <- do.call(cbind, Amat_cols)

# --- 4) solve QP
sol <- solve.QP(Dmat = Dmat, dvec = dvec, Amat = Amat, bvec = bvec, meq = meq)
w_strat <- sol$solution
names(w_strat) <- colnames(returns_matrix)

# --- 5) compute implied asset exposures and checks
implied_asset_exposure <- as.numeric(t(A_asset) %*% w_strat)  # length n_assets
names(implied_asset_exposure) <- assets

# diagnostics
sum_w <- sum(w_strat)
achieved_return <- sum(mu_s * w_strat)
port_var <- as.numeric(t(w_strat) %*% Sigma_s %*% w_strat)
min_asset_obs <- min(implied_asset_exposure)
max_asset_obs <- max(implied_asset_exposure)

list(
  w_strat = w_strat,
  implied_asset_exposure = implied_asset_exposure,
  sum_w = sum_w,
  achieved_return = achieved_return,
  port_var = port_var,
  solver_value = sol$value,
  min_asset_obs = min_asset_obs,
  max_asset_obs = max_asset_obs
)

# Use evaluation_end_date as formation date (recommended)
formation_date <- as.Date(evaluation_end_date)    # or train_end_date or whatever you intend

# Build a small weights table from the implied exposures (ensure names)
weights_from_opt <- data.frame(
  ticker = assets,
  weight = as.numeric(implied_asset_exposure),
  stringsAsFactors = FALSE
)
# if implied_asset_exposure had names, you could also do:
# weights_from_opt <- tibble(ticker = names(implied_asset_exposure), weight = as.numeric(implied_asset_exposure))

# Build full table that covers every constituent (fill missing with 0)
Weights.Portfolio <- tibble(ticker = unique(constituent_names)) %>%
  left_join(weights_from_opt, by = "ticker") %>%
  mutate(weight = ifelse(is.na(weight), 0, weight),
         Date   = formation_date) %>%
  select(Date, ticker, weight)

# Optional: normalize so the portfolio return is fractional
# If you want port_ret to be "fractional return" (i.e. starting capital = 1), normalize by sum(abs(weights))
# Otherwise the weights are notional positions, not fractions. e.g. sum abs(w) = 1.01 implies your portfolio has 101% gross exposure relative to $1 starting equation.  You are 1% levered.  You are holding $1.01 of assets financed by $1 of equity and $0.01 of debt.
normalize_weights <- TRUE   # set FALSE if you prefer not to normalize
if (normalize_weights) {
  sabs <- sum(abs(Weights.Portfolio$weight))
  if (sabs == 0) stop("All implied weights are zero — nothing to construct.")
  Weights.Portfolio <- Weights.Portfolio %>%
    mutate(weight = weight / sabs)
}

# Quick diagnostics
diag_tbl <- Weights.Portfolio %>%
  summarize(
    formation_date = first(Date),
    n_tickers = n(),
    sum_weights = sum(weight),
    sum_abs_weights = sum(abs(weight)),
    max_weight = max(weight),
    min_weight = min(weight),
    n_nonzero = sum(abs(weight) > 1e-9)
  )
print(diag_tbl)

# Show any tickers from implied exposure missing in constituent list (should be none)
missing_in_returns <- setdiff(weights_from_opt$ticker, Weights.Portfolio$ticker)
if (length(missing_in_returns) > 0) {
  message("Warning: these tickers from the optimizer are not present in the returns universe: ",
          paste(missing_in_returns, collapse = ", "))
}

# Save or inspect head
head(Weights.Portfolio)

```

## Evaluation

Task: Construct the combined portfolio and evaluate during the Test period

```{r}

Port.Combined <- construct_portfolio(
  Test.Returns.Constituents,
  Weights.Portfolio, 
  date_col   = "Date",
  ticker_col = "ticker",
  return_col = "OneDaySimpleRet",
  weight_col = "weight",
  return_type = "simple",
  rebalance_freq = "monthly",
  shift_n = 1
)

Port.Combined$portfolio
Port.Combined$holdings

```

Task: Display the strategy level weights

```{r}
Strategy_Weights <- data.frame(
  Strategy = c("Momentum", "High_SR"),
  Weight = as.numeric(w_strat)
)
Strategy_Weights
```

Task: Plot the asset level weights for the top 10 longs and shorts

```{r}
TopN <- 10
Weights.Portfolio %>%
  arrange(desc(weight)) %>%
  slice(1:TopN) %>%
  mutate(Type = "Long") %>%
  bind_rows(
    Weights.Portfolio %>%
      arrange(weight) %>%
      slice(1:TopN) %>%
      mutate(Type = "Short")
  ) %>%
  ggplot(aes(x = reorder(ticker, weight), y = weight, fill = Type)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(
    title = paste("Top", TopN, "Long and Short Asset Weights in Combined Portfolio"),
    x = "Ticker",
    y = "Weight"
  ) +
  theme_minimal()
```

Task: Plot the portfolio index with \$\$\$1 initial equity as of the last date of the evaluation period and then track that \$\$\$1 through the test period.

```{r}
port <- Port.Combined$portfolio %>% arrange(date)

anchor_date <- as.Date(evaluation_end_date)  

# create anchor row
anchor_row <- tibble(
  date = anchor_date,
  gross_exposure = port$gross_exposure[1],  # optional: or set to NA/0
  pnl = 0,
  port_ret = 0,
  port_index = 1,
  n_assets = port$n_assets[1],
  n_hold = port$n_hold[1],
  n_long = port$n_long[1],
  n_short = port$n_short[1]
)

# ensure anchor is strictly before first test-day (or equal if you want)
if (anchor_date >= min(port$date)) {
  anchor_row$date <- min(port$date) - 1
}

port_with_anchor <- bind_rows(anchor_row, port) %>% arrange(date)

# plot
ggplot(port_with_anchor, aes(x = date, y = port_index)) +
  geom_line(color = "blue") +
  labs(title = "Portfolio Index During the Test Period",
       x = "",
       y = "Portfolio Index value of $1 invested at end of evalulation period") +
  theme_minimal()
```
