---
title: "Momentum Strategy"
subtitle: "Constructing and evaluating a momentum strategy"
author: Mike Aguilar | https://www.linkedin.com/in/mike-aguilar-econ/ 
format: html
editor: visual
toc: true
toc-depth: 5
toc-location: left
embed-resources: true
execute: 
  warning: false
  echo: true
---

# Housekeeping

```{r}
#| echo: false
cat("\014")  # clear console
rm(list=ls())  # Clear the workspace
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
suppressPackageStartupMessages(library(here))
source(here("Supporting", "PackageLoads.R"))

```

# Data

## Load monthly momentum from GitHub

```{r}
Momentum <- ?(here("Data","SP500_Mthly_Momentum.csv"))
```

## Set Test and Train Periods

Task: Create test_date as 9/30/2025 and compute the last train date as one month prior

```{r}
test_date <- as.Date("2025-09-30") 
last_train_date <- as.Date(
  format(test_date, "%Y-%m-01")  ) - 1
```

Task: Create Train.Momentum.Constituents which contains all the dates in the traning period only for the SP500 constituents

```{r}
Train.Momentum.Constituents <- Momentum %>%
  filter(Date <= last_train_date) %>%
  filter(!ticker == ?)
```

# Signal Construction

Task: Construct a simple momentum strategy

-   Define LongCondition = 1 if $mom^{Train}_{i} >0$ and 0 otherwise

-   Define ShortCondition = 1 if $mom^{Train}_{i} < 0$ and 0 otherwise

-   $s_{i,t}$ = LongCondition minus ShortCondition

Task: Create a vector that contains the momentum on the last observation of the training period

```{r}

Signal_Matrix <- ? %>% 
  filter(Date == last_train_date) %>%
  select(Date, ticker, momentum) %>% 
  rename("signal_matrix" = "momentum")

Signal_Matrix
```

# Daily Data for Evaluation

Task: Load SP500 Daily Simple Returns

```{r}
SP500_Daily_SimpleReturns<-read.csv(?("Data","SP500_Daily_SimpleReturns.csv"))
```

Task: ensure Date is Date class

```{r}
SP500_Daily_SimpleReturns <- SP500_Daily_SimpleReturns %>%
  mutate(Date = as.Date(Date))
```

Task: Filter test period (dates AFTER last_train)

```{r}
test_returns <- SP500_Daily_SimpleReturns %>%
  filter(
    Date ? as.Date(last_train_date) &
    Date ? as.Date(test_date)
  ) 
```

# Raw Weights

Task: Create a weights dataframe based upon the signal. Set the weight equal to the signal value.

```{r}
weight_df <- Signal_Matrix %>% 
  transmute(
    ticker,
    Date,
    weight = signal_matrix) 
weight_df

```

Task: Construct the portfolio

```{r}
Port.Raw <- construct_portfolio(
    ?,
    ?,
    # column names in inputs
    date_col   = "Date",
    ticker_col = "ticker",
    return_col = "OneDaySimpleRet",
    weight_col = "weight",
    return_type = "simple", #c("simple","log"),
    rebalance_freq = "monthly", #c("monthly","quarterly","yearly"),
    shift_n = 1 # if shift_n=1 and rebalance_freq=monthly, then w_Jan * ret_Feb
)

```

Task: Plot the port_index over time

```{r}
Port.Raw$portfolio %>%
  ggplot(aes(x = date, y = ?)) +
  geom_line(color = "blue") +
  labs(title = "Portfolio Index Over Time",
       x = "",
       y = "Portfolio Index") +
  theme_minimal()
```

Task: Display the portfolio holdings

```{r}
head(Port.Raw$?, 50)
```

Task: Display the portfolio time series

```{r}
Port.Raw$?
```

Q: What does gross exposure mean?

A:



Q: Why can't you use that short sale cash to reduce the amount you borrow in practice? 

A: 

Q: Contrast the meaning of gross exposure, pnl, port_ret, port index. 

A: 




# Fully Invested Weights



Task: Compute the gross_scale as the sum of the abs value of the signals
```{r}
gross_scale <- ?(abs(Signal_Matrix$signal_matrix))
gross_scale
```
Q: What is the meaning of gross_scale in our strategy context? Why different from gross_exposure? 

A: 



Task: Create fully invested weights dataframe by dividing the signal by gross_scale
```{r}
fullyinvested_weights_df <- weight_df %>%
  mutate(weight = Signal_Matrix$signal_matrix ?)

sum(abs(fullyinvested_weights_df$weight)) 
```
Q: What's the intuition behind constructing the strategy weights this way? 

A: 



Task: Construct the portfolio using construct_portfolio

```{r}
Port.FullyInvested <- construct_portfolio(
    ?,
    ?,
    # column names in inputs
    date_col   = "Date",
    ticker_col = "ticker",
    return_col = "OneDaySimpleRet",
    weight_col = "weight",
    return_type = "simple", #c("simple","log"),
    rebalance_freq = "monthly", #c("monthly","quarterly","yearly"),
    shift_n = 1 # if shift_n=1 and rebalance_freq=monthly, then w_Jan * ret_Feb
)

```

Task: Plot the port_index over time

```{r}
Port.FullyInvested$portfolio %>%
  ggplot(aes(x = date, y = ?)) +
  geom_line(color = "blue") +
  labs(title = "Portfolio Index Over Time",
       x = "",
       y = "Portfolio Index") +
  theme_minimal()
```

Task: Create a plot that overlays the two portfolio indices

```{r}
combined_portfolio <- Port.Raw$portfolio %>%
  select(date, raw_port_index = port_index) %>%
  inner_join(
    Port.FullyInvested$portfolio %>%
      select(date, fully_invested_port_index = port_index),
    by = "date"
  )
combined_portfolio %>%
  ggplot(aes(x = date)) +
  geom_line(aes(y = raw_port_index, color = "Raw Weights")) +
  geom_line(aes(y = ?, color = "Fully Invested Weights")) +
  labs(title = "Portfolio Index Comparison",
       x = "",
       y = "Portfolio Index") +
  scale_color_manual(values = c("Raw Weights" = "blue", "Fully Invested Weights" = "red")) +
  theme_minimal()
```
Q: Why is there such a large difference between these seemingly similar strategies?

A: 

