---
title: "Momentum Signal"
subtitle: "Constructing and evaluating a static momentum signal"
author: Mike Aguilar | https://www.linkedin.com/in/mike-aguilar-econ/ 
format: html
editor: visual
toc: true
toc-depth: 5
toc-location: left
embed-resources: true
execute: 
  warning: false
  echo: true
---

# Housekeeping

```{r}
#| echo: false
cat("\014")  # clear console
rm(list=ls())  # Clear the workspace
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
suppressPackageStartupMessages(library(here))
?

```

# Data

## Load simple returns

Task: Load SP500-Monthly Simple Returns

```{r}
Returns <- ?(here("Data","SP500_Mthly_SimpleReturns.csv"))
```

## Load monthly momentum from GitHub

```{r}
Momentum <- read.csv(?("Data","SP500_Mthly_Momentum.csv"))
```

## Merge

Task: Merge the returns and momentum, retaining only those dates with complete information

```{r}
merged_df <- Returns %>%
  left_join(Momentum, ? = c("Date", "ticker"))

```

## Set Test and Train Periods

Task: Create a dataframe for the Train period which spans the beginning of the dataset to Aug2025. Call this dataframe Test. Create a dataframe for the Train period which is Sep2025 period. Call this dataframe Train.

```{r}

test_date <- as.Date("2025-09-30") 
last_train_date <- as.Date(
  format(test_date, "%Y-%m-01")  ) - 1

Train <- merged_df ? filter(Date < test_date)
Train.Returns  <- Train %>% select(!momentum)
Train.Momentum <- Train %>% select(!OneMthSimpleRet)

Test <- merged_df %>% ?(Date == test_date)
Test.Returns  <- Test %>% select(!momentum)
Test.Momentum <- Test %>% select(!OneMthSimpleRet)

```

## Isolate Index and Constituents

Task: Isolate the returns for the index and the group of constituents into separate dataframes. Repeat for momentum.

```{r}
Train.Constituents <- Train %>%
  filter(? == "SP500")
Train.Returns.Index <- Train.Returns %>%
  filter(ticker == "SP500")
Train.Returns.Constituents <- Train.Returns %>%
  filter(!ticker == "SP500")
Train.Momentum.Index <- Train.Momentum %>%
  filter(ticker == "SP500")
Train.Momentum.Constituents <- Train.Momentum %>%
  filter(!ticker == "SP500")

Test.Constituents <- Test %>%
  filter(!ticker == "SP500")
Test.Returns.Index <- Test.Returns %>%
  filter(ticker == "SP500")
Test.Returns.Constituents <- Test.Returns %>%
  filter(!ticker == "SP500")
Test.Momentum.Index <- Test.Momentum %>%
  filter(ticker == "SP500")
Test.Momentum.Constituents <- Test.Momentum %>%
  filter(!ticker == "SP500")

```

# Signal Construction

Task: Construct a simple momentum strategy

-   Define LongCondition = 1 if $mom^{Train}_{i} >0$ and 0 otherwise

-   Define ShortCondition = 1 if $mom^{Train}_{i} < 0$ and 0 otherwise

-   $s_{i,t}$ = LongCondition minus ShortCondition

Task: Create a dataframe that contains the momentum on the last observation of the training period

```{r}

Signal_Matrix <- ? %>% 
  filter(Date == last_train_date) %>%
  select(ticker, momentum) %>% 
  rename("signal_matrix" = "momentum")

Signal_Matrix
```

Task: Create a dataframe that takes +1,0,-1 based upon long, neutral, short positioning. Long of signal \> tau. Short if signal less than -tau. Neutral otherwise.

```{r}

tau <- 0

Signal_Position <- Signal_Matrix %>%
  transmute(
    ticker,
    signal = ?(
      signal_matrix >  tau  ~  1L,
      signal_matrix < -tau  ~ -1L,
      TRUE             ~  0L
    )
  ) %>%
  rename("signal_position" = "signal")

Signal_Position

```

# Evaluate

Task: Format Test period returns to conform to the Signal Testing function

```{r}
test_returns <- Test.Returns.Constituents %>%
  ?(ticker,OneMthSimpleRet) %>%
  rename(test_returns = OneMthSimpleRet)
```

Task: Create meta labels

```{r}
 Meta <- list(
   assetname = "asset",
   benchmarkname = "benchmark",
   signalname = "Momentum"
 )
```

Task: Implement the Signal Evaluation Static function

```{r}
return_threshold = 0

  SignalEval <- ?(
    test_returns = test_returns,
    Meta = Meta,
    return_threshold = return_threshold,
    signal_position = Signal_Position
  )
```

Task: Display and interpret the Trade results

```{r}
SignalEval$TradeResults_Percent
```

A:

Task: Display and interpret the hit table

```{r}
SignalEval$hit_table
```

A:

Task: Display and interpret the Info Coeffient

```{r}
SignalEval$summary_ic_tbl
```

A:

Task: Display and interpret the confusion matrix. Focus on the Recall, Specificity, and Type 1 error for Long and Short.

```{r}
SignalEval$confusion_report
```

A:
