---
title: "SP500 EDA"
subtitle: "Exploring SP500 returns"
author: Mike Aguilar | https://www.linkedin.com/in/mike-aguilar-econ/ 
format: html
editor: visual
toc: true
toc-depth: 5
toc-location: left
embed-resources: true
execute: 
  warning: false
  echo: true
---

# Background

In one of the primary examples for this course we will be working with data on the SP500.

Our empirical goal here is to conduct some Exploratory Descriptive Analysis (EDA).

# Housekeeping

```{r}
#| echo: false

cat("\014")  # clear console
rm(list=ls())  # Clear the workspace
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
suppressPackageStartupMessages(library(here))
source(here("Supporting", "PackageLoads.R"))

```

# Data Acquisition

## Load simple returns

Task: Load SP500-Monthly-SimpleReturns 

```{r}

SP500_Mthly_SimpleReturns<- read.csv(here("Data","SP500_Mthly_SimpleReturns.csv"))
```

## Set Test and Train Periods

Task: Create a dataframe for the Train period which spans the beginning of the dataset to Aug2025. Call this dataframe Test. Create a dataframe for the Train period which is Sep2025 period. Call this dataframe Train.

```{r}

test_date <- as.Date("2025-09-30") 
last_train_date <- as.Date(
  format(test_date, "%Y-%m-01")  ) - 1

Train <- SP500_Mthly_SimpleReturns %>% filter(Date < test_date)


Test <- SP500_Mthly_SimpleReturns %>% filter(Date == test_date)


```

Task: Separate the Index and Constituents

```{r}
Train.Constituents <- Train %>%
  filter(!ticker == "SP500")


Train.Index <- Train %>%
  filter(ticker == "SP500")

Test.Constituents <- Test %>%
  filter(!ticker == "SP500" )

Test.Index <- Test %>%
  filter(ticker == "SP500")

```

# EDA

## Training Period

Assume that all mentions of "returns" in this subsection refer to Train.

### Average Returns

Task: Compute mean 1mth returns for all of the assets

```{r}
Mean_1m_Returns_Train <- Train.Constituents %>%
  group_by(ticker) %>%
  summarise(mean_ret_1m = mean(OneMthSimpleRet, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_ret_1m))

Mean_1m_Returns_Train
```

Task: Create a relative frequency histogram with Gaussian overlay

```{r}
x <- Train.Constituents$OneMthSimpleRet

mu <- mean(x)
sig <- sd(x)

df <- data.frame(ret_1m = x)

ggplot(df, aes(x = ret_1m)) +
  # Histogram as density
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50, fill = "white", color = "steelblue") +
  # Shade the fitted Gaussian density
  stat_function(fun = dnorm,
                args = list(mean = mu, sd = sig),
                geom = "area",
                fill = "steelblue",
                alpha = 0.20) +
  # Gaussian outline on top
  stat_function(fun = dnorm,
                args = list(mean = mu, sd = sig),
                linewidth = 1,
                color = "steelblue") +
  labs(x = "Per-asset Avg of 1-month simple return", y = "Density")
```

Task: Create a table of sample stats including: mean,sigma, skew, kurtosis, min, q.1, q.25, q.5, q.75, q.9, max

```{r}
ret_stats <- tibble(
  mean  = mean(x, na.rm = TRUE),
  sigma = sd(x, na.rm = TRUE),
  skew  = moments::skewness(x, na.rm = TRUE),
  kurt  = moments::kurtosis(x, na.rm = TRUE),
  min   = min(x, na.rm = TRUE),
  q_01  = unname(quantile(x, 0.01, na.rm = TRUE)),
  q_10  = unname(quantile(x, 0.10, na.rm = TRUE)),
  q_25  = unname(quantile(x, 0.25, na.rm = TRUE)),
  q_50  = unname(quantile(x, 0.50, na.rm = TRUE)),
  q_75  = unname(quantile(x, 0.75, na.rm = TRUE)),
  q_90  = unname(quantile(x, 0.90, na.rm = TRUE)),
  max   = max(x, na.rm = TRUE)
)


ret_stats%>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "stat",
    values_to = "value"
  ) %>%
  dplyr::mutate(value = round(value, 3))


```

### Sigma

Task: Compute the Standard Deviation of returns for all the assets

```{r}
SD_1m_Returns_ByAsset <- Train.Constituents %>%
  group_by(ticker) %>%
  summarise(sd_ret_1m = sd(OneMthSimpleRet, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(sd_ret_1m))

SD_1m_Returns_ByAsset
```

Task: Create a relative frequency histogram with Gaussian overlay

```{r}
x <- SD_1m_Returns_ByAsset$sd_ret_1m

mu <- mean(x)
sig <- sd(x)

df <- data.frame(ret_1m = x)

ggplot(df, aes(x = ret_1m)) +
  # Histogram as density
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50, fill = "white", color = "steelblue") +
  # Shade the fitted Gaussian density
  stat_function(fun = dnorm,
                args = list(mean = mu, sd = sig),
                geom = "area",
                fill = "steelblue",
                alpha = 0.20) +
  # Gaussian outline on top
  stat_function(fun = dnorm,
                args = list(mean = mu, sd = sig),
                linewidth = 1,
                color = "steelblue") +
  labs(x = "Standard deviaiton of per-asset average 1-month simple return", y = "Density")
```

Task: Create a table of sample stats including: mean,sigma, skew, kurtosis, min, q.1, q.25, q.5, q.75, q.9, max

```{r}
sd_stats <- tibble(
  mean  = mean(x, na.rm = TRUE),
  sigma = sd(x, na.rm = TRUE),
  skew  = moments::skewness(x, na.rm = TRUE),
  kurt  = moments::kurtosis(x, na.rm = TRUE),
  min   = min(x, na.rm = TRUE),
  q_01  = unname(quantile(x, 0.01, na.rm = TRUE)),
  q_10  = unname(quantile(x, 0.10, na.rm = TRUE)),
  q_25  = unname(quantile(x, 0.25, na.rm = TRUE)),
  q_50  = unname(quantile(x, 0.50, na.rm = TRUE)),
  q_75  = unname(quantile(x, 0.75, na.rm = TRUE)),
  q_90  = unname(quantile(x, 0.90, na.rm = TRUE)),
  max   = max(x, na.rm = TRUE)
)


sd_stats %>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "stat",
    values_to = "value"
  ) %>%
  dplyr::mutate(value = round(value, 3))
```

### Skew

Task: Compute the Skewness of returns for all the assets

```{r}
Skew_1m_Returns_ByAsset <- Train.Constituents %>%
  group_by(ticker) %>%
  summarise(skew_ret_1m = moments::skewness(OneMthSimpleRet, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(skew_ret_1m))

Skew_1m_Returns_ByAsset
```

Task: Create a relative frequency histogram with Gaussian overlay

```{r}
x <- Skew_1m_Returns_ByAsset$skew_ret_1m

mu <- mean(x)
sig <- sd(x)

df <- data.frame(ret_1m = x)

ggplot(df, aes(x = ret_1m)) +
  # Histogram as density
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50, fill = "white", color = "steelblue") +
  # Shade the fitted Gaussian density
  stat_function(fun = dnorm,
                args = list(mean = mu, sd = sig),
                geom = "area",
                fill = "steelblue",
                alpha = 0.20) +
  # Gaussian outline on top
  stat_function(fun = dnorm,
                args = list(mean = mu, sd = sig),
                linewidth = 1,
                color = "steelblue") +
  labs(x = "Skewness of of per-asset average 1-month simple return", y = "Density")
```

Task: Create a table of sample stats including: mean,sigma, skew, kurtosis, min, q.1, q.25, q.5, q.75, q.9, max

```{r}
skew_stats <- tibble(
  mean  = mean(x, na.rm = TRUE),
  sigma = sd(x, na.rm = TRUE),
  skew  = moments::skewness(x, na.rm = TRUE),
  kurt  = moments::kurtosis(x, na.rm = TRUE),
  min   = min(x, na.rm = TRUE),
  q_01  = unname(quantile(x, 0.01, na.rm = TRUE)),
  q_10  = unname(quantile(x, 0.10, na.rm = TRUE)),
  q_25  = unname(quantile(x, 0.25, na.rm = TRUE)),
  q_50  = unname(quantile(x, 0.50, na.rm = TRUE)),
  q_75  = unname(quantile(x, 0.75, na.rm = TRUE)),
  q_90  = unname(quantile(x, 0.90, na.rm = TRUE)),
  max   = max(x, na.rm = TRUE)
)


skew_stats %>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "stat",
    values_to = "value"
  ) %>%
  dplyr::mutate(value = round(value, 3))
```

### Kurtosis

Task: Compute the Kurtosis of returns for all the assets

```{r}
Kurt_1m_Returns_ByAsset <- Train.Constituents %>%
  group_by(ticker) %>%
  summarise(kurt_ret_1m = moments::kurtosis(OneMthSimpleRet, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(kurt_ret_1m))

Kurt_1m_Returns_ByAsset
```

Task: Create a relative frequency histogram with Gaussian overlay

```{r}
x <- Kurt_1m_Returns_ByAsset$kurt_ret_1m
  
mu <- mean(x)
sig <- sd(x)

df <- data.frame(ret_1m = x)

ggplot(df, aes(x = ret_1m)) +
  # Histogram as density
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50, fill = "white", color = "steelblue") +
  # Shade the fitted Gaussian density
  stat_function(fun = dnorm,
                args = list(mean = mu, sd = sig),
                geom = "area",
                fill = "steelblue",
                alpha = 0.20) +
  # Gaussian outline on top
  stat_function(fun = dnorm,
                args = list(mean = mu, sd = sig),
                linewidth = 1,
                color = "steelblue") +
  labs(x = "Kurtosis of per-asset average 1-month simple return", y = "Density")
```

Task: Create a table of sample stats including: mean,sigma, skew, kurtosis, min, q.1, q.25, q.5, q.75, q.9, max

```{r}
kurt_stats <- tibble(
  mean  = mean(x, na.rm = TRUE),
  sigma = sd(x, na.rm = TRUE),
  skew  = moments::skewness(x, na.rm = TRUE),
  kurt  = moments::kurtosis(x, na.rm = TRUE),
  min   = min(x, na.rm = TRUE),
  q_01  = unname(quantile(x, 0.01, na.rm = TRUE)),
  q_10  = unname(quantile(x, 0.10, na.rm = TRUE)),
  q_25  = unname(quantile(x, 0.25, na.rm = TRUE)),
  q_50  = unname(quantile(x, 0.50, na.rm = TRUE)),
  q_75  = unname(quantile(x, 0.75, na.rm = TRUE)),
  q_90  = unname(quantile(x, 0.90, na.rm = TRUE)),
  max   = max(x, na.rm = TRUE)
)


kurt_stats %>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "stat",
    values_to = "value"
  ) %>%
  dplyr::mutate(value = round(value, 3))
```

## Testing Period

Assume that all mentions of "returns" in this subsection refer to Test

### Summarize

Task: Create a histogram of returns

```{r}
x <- Test.Constituents$OneMthSimpleRet 

mu <- mean(x)
sig <- sd(x)

df <- data.frame(ret_1m = x)

ggplot(df, aes(x = ret_1m)) +
  # Histogram as density
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50, fill = "white", color = "steelblue") +
  # Shade the fitted Gaussian density
  stat_function(fun = dnorm,
                args = list(mean = mu, sd = sig),
                geom = "area",
                fill = "steelblue",
                alpha = 0.20) +
  # Gaussian outline on top
  stat_function(fun = dnorm,
                args = list(mean = mu, sd = sig),
                linewidth = 1,
                color = "steelblue") +
  labs(x = "Per-asset average 1-month simple return (Test)", y = "Density")
```

Task: Create a table of sample stats including: mean,sigma, skew, kurtosis, min, q.1, q.25, q.5, q.75, q.9, max

```{r}
ret_stats_test <- tibble(
  mean  = mean(x, na.rm = TRUE),
  sigma = sd(x, na.rm = TRUE),
  skew  = moments::skewness(x, na.rm = TRUE),
  kurt  = moments::kurtosis(x, na.rm = TRUE),
  min   = min(x, na.rm = TRUE),
  q_01  = unname(quantile(x, 0.01, na.rm = TRUE)),
  q_10  = unname(quantile(x, 0.10, na.rm = TRUE)),
  q_25  = unname(quantile(x, 0.25, na.rm = TRUE)),
  q_50  = unname(quantile(x, 0.50, na.rm = TRUE)),
  q_75  = unname(quantile(x, 0.75, na.rm = TRUE)),
  q_90  = unname(quantile(x, 0.90, na.rm = TRUE)),
  max   = max(x, na.rm = TRUE)
)


ret_stats_test %>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "stat",
    values_to = "value"
  ) %>%
  dplyr::mutate(value = round(value, 3))
```

## Test / Train Similarity

Task: Create a scatter plot with Test on the vertical and per-asset average return in the Train period on the horizontal.

```{r}

# 1) Per-asset average return in Train (x-axis)
train_mean <- Train.Constituents %>%
  mutate(ret_1m = as.numeric(OneMthSimpleRet)) %>%
  group_by(ticker) %>%
  summarise(train_mean_ret = mean(ret_1m, na.rm = TRUE), .groups = "drop")

# 2) Per-asset return in Test (y-axis)
test <- Test.Constituents %>%
  mutate(ret_1m = as.numeric(OneMthSimpleRet)) %>%
  group_by(ticker) %>%
  summarise(test_ret = ret_1m, .groups = "drop")

# 3) Merge + scatter
df <- train_mean %>%
  inner_join(test, by = "ticker")

ggplot(df, aes(x = train_mean_ret, y = test_ret)) +
  geom_point(alpha = 0.7, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "darkred") +
  labs(
    x = "Per-asset average 1-month return (Train)",
    y = "Per-asset average 1-month return (Test)"
  )
```

Task: Compute the correlation between Testing period return and the per-asset average return in the Training period.

```{r}
correlation <- cor(df$train_mean_ret, df$test_ret, use = "complete.obs")
correlation
```

Task: Compute the ranks of each of the returns in the Testing period

```{r}
# Rank each monthly return across assets within each date in the Test period
# (1 = lowest return, larger = higher return). Use ties.method="average" by default
test_ranked <- test %>%
  mutate(
    test_rank = rank(test_ret, na.last = "keep", ties.method = "average")
   ) %>%
  ungroup() %>%
  arrange((test_ret))

test_ranked
```

Task: Compute the ranks of each of the per-asset average returns in the Training period

```{r}
train_ranked <- train_mean %>%
  mutate(
    train_rank = rank(train_mean_ret, na.last = "keep", ties.method = "average")
  ) %>%
  arrange(train_rank)

train_ranked
```

Task: Compute the correlation among the ranks

```{r}
rank_df <- train_ranked %>%
  inner_join(test_ranked, by = "ticker") 

rank_corr <- cor(rank_df$train_rank, rank_df$test_rank, use = "complete.obs", method = "pearson")
rank_corr
```
